---
title: "Reading Multiple csvs as 1 data frame"
subtitle: "Reading chunked csv files"
date: '2018-10-27'
slug: read-chunked-csv
categories:
  - R
  - tutorial
tags:
  - R 
  - tidyverse
  - tutorial
  - data manipulation
---


In an earlier [posting](josiahparry.com/post/write-chunked-csv) I wrote about having to break a single csv into multiple csvs. In other scenarios one data set maybe provided as multiple a csvs. 

Thankfully `purrr` has a beautiful function called `map_df()` which will make this into a two liner. This process has essentially 3 steps.

1. Create a vector of all `.csv` files that should be merged together.
2. Read each file using `readr::read_csv()`
3. Combine each dataframe into one.

`map_df()` maps (applys) a function to each value of an object and produces a dataframe of all outputs. 

For this example I will use the csvs I created in a previous tutorial utilizing a dataset from the Quantitative Social Science book. 

```{r}
# Get all csv file names 
file_names <- list.files("../../static/data/chunk_data", pattern = "\\.csv", full.names = TRUE)
file_names
```

```{r}
library(tidyverse)
# apply 
all_csvs <- map_df(file_names, read_csv)

# preview the data
head(all_csvs)
```
